Staged Abstract Interpreters

### 2019-09-06 paper/50_stageabs.tex
typo

Our staging schema works by unfolding the interpreter over the input AST. In
practice, however, it would suffer from code explosion when analyzing
(specializing) large programs, which increases the compile time. If we generate
next stage programs running on JVM, such large generated program would also
potentially cause runtime GC overhead. In this section, we present several
optimizations that largely mitigate these issues. For all that, implementing
these optimizations do not need to change the generic interpreter.

@@

Although every component except @Expr@s are staged, we treat the data
structures such as @Map@s as black-boxes, which means the operations on a @Map@
directly become code in the next stage, but we don't inspect any further inside.
As we identified when introducing the generic interface, the keys of @Env@s are
identifiers (i.e., strings) in the program, which are completely known
statically. This leaves us a chance to further specialize the data structures.
For example, let us assume that the @Map[K, V]@ is implemented as a hash map. If
the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values; the size of the array is
also known statically. An access to the map is translated into an access to the
array with pre-determined index during staging.

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the set of identifiers in the program. Then the accesses to the
environment can be completely determined statically and generates the addresses
directly. After which, the stores can be specialized as arrays of @Rep[Value]@
elements.

\paragraph{Selective Caching} It is observed that the two-fold co-inductive
caching is used for every recursive call of our abstract interpreter. But this
is not necessary and even very expensive when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
\cite{Flanagan:1993:ECC:155090.155113}, we may use a selective caching
algorithm that does not generate caching for atomic expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

\paragraph{Partially-static Data}

Our treatment to binding-times is coarse-grained: @Expr@s is static, the rest of
the world are all dynamic. But this is not always true since the static data
has to be used somewhere with the dynamic operation. Partial-static data is a
way to improve binding-times and optimize the generated code.

For example, to fold a singleton set (often appears in @SetT@), e.g.,
@Set(x).foldLeft(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully apply @foldLeft@ to the set with an anonymous
function. But we can also utilize the algebraic property of @foldLeft@ to
generate cheaper code: @Set(x).foldLeft(init)(f) = f(init, x)@. Since the
function @f@ is known at the current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partial-static data structures, such as @Set@ and @Map@, which greatly reduces
the size of residual programs.

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an \textit{operational} perspective.
In this section, we review and summarize our recipe to achieve the staged
abstract interpreter and discuss correctness issues and different design choices.

@@

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding times,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item By using an appropriate monad stack, we can implement a modular, unstaged,
    abstract interpreter. This step has been explored in the previous literature.
  \item Replacing the monad stack to a staged monad stack; such monads operate on
    staged data types, i.e., next-stage values. We also need to reimplement several
    primitive operations on staged values, in a modular fashion.
\end{itemize}

Monadic interpreters are known to be able to decouple the interpretation process
and the underlying semantics. The key insight in this paper is that by making the
monadic interpreter binding-time polymorphic, it can be extended for generating
efficient code in a staging framework. The underlying semantics and staging are
two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged: only data (such as sets and maps) are
staged. All the monadic computation, i.e., functions passed to the monadic bind
operation @flatMap@, are statically known. This is why we can eliminate the
monadic layer and its associated overhead.

\paragraph{What has been eliminated?} In the generated code, all the
primitive operations (such as @eval@, @fix@, @ap_clo@, etc.) and monadic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representations are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the use of monads internally.

@@

Soundness is the central concern of static analysis, and as such, is vital
for prospective users. We now briefly examine how our
approach preserves the correctness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here is based on empirical evidence; a rigorous proof for soundness
preservation of the staged abstract interpreters remains an open question.

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistence, i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal Form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularly.
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structures we used here are simply
    black boxes that wrap the Scala's library data structures.
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

@@

We implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizing Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering effort.

>>>>

Our staging schema works by unfolding the interpreter over the input abstract
syntax tree. In practice, however, it would suffer from code explosion when
analyzing (specializing) large programs, which increases the compile time. If
we generate next-stage programs running on JVM, such large generated program
would also potentially cause runtime GC overhead. In this section, we present
optimizations that largely mitigate these issues. For all that,
implementing these optimizations do not need to change the generic interpreter.

@@

Every components in the interpreter except @Expr@s are staged. The data
structures representing these components are treated as black-boxes, such as
@Map@, which means that the operations on a @Map@ directly become nex-stage code,
but we don't inspect any further inside. As we identified when
introducing the generic interface, the keys of @Env@s are identifiers (i.e.,
strings) in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures.  For
example, let us assume that the @Map[K, V]@ is implemented as a hash map. If
all the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus, the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values, and the size of the array is
known statically. An access to the map is translated into an access to the
array with the pre-computed index during staging.

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the set of identifiers in the program. Then the accesses to
the environment can be completely computed statically and we may generate the
addresses directly during staging. After which, the stores are specialized
as arrays with element type @Rep[Value]@.

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive calls in our abstract interpreter. But this
is not necessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
$\lambda$-calculus \cite{Flanagan:1993:ECC:155090.155113}, we can use a
selective caching algorithm that does not generate caching code for atomic
expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

\paragraph{Partially-static Data}
Our treatment of binding-times is coarse-grained: @Expr@s is static, the rest
of the components are all dynamic. But this is not always true since the static
data has to be used somewhere with dynamic operations. Partial-static data is a
way to improve binding-times and optimize the generated code.

For example, to fold a static singleton set (often appears in @SetT@), e.g.,
@Set(x).foldLeft(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully generates code that applies @foldLeft@ with the set and
function @f@. But we can also utilize the algebraic property of @foldLeft@ to
generate cheaper code: @Set(x).foldLeft(init)(f) = f(init, x)@. Since the
function @f@ is known at the current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partial-static patterns, such as for @Set@ and @Map@. This optimizations
greatly reduces the size of residual programs.

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an operational perspective.  In this section, we
review and summarize our recipe to achieve the staged abstract interpreter and
discuss correctness issues and different design choices.

@@

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding times,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item Then we can implement an unstaged abstract interpreter modularly using
    the appropriate monad stack. This step has been explored in the previous
    literatures.
  \item Finally, we replace the monad stack with a staged monad stack, and
    refactor related primitive operations.  Such staged monads operate on
    staged data types, i.e., next-stage values.
\end{itemize}

Monadic interpreters are known to be able to decouple the interpretation
procedure and the underlying semantics. The key insight in this paper is that
by making the monadic interpreter binding-time polymorphic, the interpreter can
be extended for generating efficient code. The underlying semantics and staging
are two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged: only data (such as sets and maps)
are staged. All the monadic computation, i.e., functions passed to the monadic
bind operation @flatMap@, are statically known. This is why we can eliminate
the monadic layer and its associated overhead.

\paragraph{What has been eliminated?} In the generated code, all the
primitive operations (such as @eval@, @fix@, @ap_clo@, etc.) and monadic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representations are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the internal code fragments of monads.

@@

Soundness is the central concern of static analysis, and as such, is vital
for prospective users. We now briefly examine how our
approach preserves the correctness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here is based on empirical evidence; a rigorous proof for soundness
preservation of the staged abstract interpreter remains an open challenge.

\begin{itemize}
  \item We usually assume the underlying MSP system preserves the meaning during staging,
    i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous sections, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of the implementation of staged monads modularly.
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    versions. For instance, the staged data structures we use here are simply
    black boxes that wrap the data structures in Scala library.
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

@@

We implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive calls of our abstract
interpreter is applied to the proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize a small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} presented for
optimizing Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering effort.
